---
title: "Homeless Demographic, race and age"
author: "Maya Fujita"
date: "2024-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
getwd()
```

###. Packages I might use
```{r}
library(tidyverse)
library(dplyr)
library(janitor)
library(ggplot2)
library(shiny)
library(shinydashboard)
library(palmerpenguins)
library(naniar)
```
```{r}
#install.packages("shinydashboard")
```
###. Loading Data
```{r}
race <- read_csv("race.csv",na = c("NA", " ", "*"))
race_homeless<- clean_names(race)
```
```{r}
age <- read_csv("age.csv")
age <- clean_names(age)
```

###. Exploring Data
```{r}
glimpse(race_homeless)
```

```{r}
glimpse(age)
```
```{r}
miss_var_summary(race_homeless)
```
###. Mean Homeless count per year (table)
```{r}
race_homeless %>% 
  mutate(calendar_year=as.factor(calendar_year)) %>%
  mutate(experiencing_homelessness_cnt=as.numeric(experiencing_homelessness_cnt)) %>% 
  group_by(calendar_year) %>% 
  filter(experiencing_homelessness_cnt!="NA") %>% 
  summarize(mean_num_homeless=mean(experiencing_homelessness_cnt))

```
###. Mean Homeless count per year (graph)
```{r}
race_homeless %>% 
  mutate(calendar_year=as.factor(calendar_year)) %>%
  mutate(experiencing_homelessness_cnt=as.numeric(experiencing_homelessness_cnt)) %>% 
  group_by(calendar_year) %>% 
  filter(experiencing_homelessness_cnt!="NA") %>% 
  summarize(mean_num_homeless=mean(experiencing_homelessness_cnt)) %>% 
  ggplot(aes(x=calendar_year, y=mean_num_homeless))+
  geom_col(color="black", fill="lightblue", alpha=0.75)+
  labs(title = "Number of Homeless People Recieving Aid from 2017-2023",
       x="Year",
       y="Average Number of Homeless Individuals")+
  geom_smooth(method = lm, se=T, na.rm = T)

```
###. Average number of homeless individuals per year after covid began
```{r}
race_homeless %>% 
  mutate(calendar_year=as.factor(calendar_year)) %>%
  mutate(experiencing_homelessness_cnt=as.numeric(experiencing_homelessness_cnt)) %>% 
  filter(calendar_year=="2020"| calendar_year=="2021"|calendar_year=="2022"|calendar_year=="2023") %>% 
  filter(experiencing_homelessness_cnt!="NA") %>% 
  summarise(mean_covid_yrs=mean(experiencing_homelessness_cnt))

```
###. Average number of homeless individuals per year before covid began
```{r}
race_homeless %>% 
  mutate(calendar_year=as.factor(calendar_year)) %>%
  mutate(experiencing_homelessness_cnt=as.numeric(experiencing_homelessness_cnt)) %>% 
  filter(calendar_year=="2017"| calendar_year=="2018"|calendar_year=="2019") %>% 
  filter(experiencing_homelessness_cnt!="NA") %>% 
  summarise(mean_non_covid_yrs=mean(experiencing_homelessness_cnt))
```
### Overall by race
```{r}
race_homeless %>% 
  
  mutate(experiencing_homelessness_cnt=as.numeric(experiencing_homelessness_cnt)) %>% 
  filter(experiencing_homelessness_cnt!="NA") %>%
  group_by(race) %>% 
  summarize(mean_count=mean(experiencing_homelessness_cnt)) %>% 
  arrange(-mean_count)
  
  

```
```{r}
view(race_homeless)
```
### Race by yolo county
```{r}
race_homeless %>% 
  mutate(experiencing_homelessness_cnt=as.numeric(experiencing_homelessness_cnt)) %>% 
  filter(experiencing_homelessness_cnt!="NA") %>%
  filter(location=="Yolo County CoC") %>% 
  group_by(race) %>% 
  summarize(mean_count=mean(experiencing_homelessness_cnt)) %>% 
  arrange(-mean_count)
```
```{r}
race_homeless %>% 
  filter(location=="Yolo County CoC") %>% 
  mutate(experiencing_homelessness_cnt=as.numeric(experiencing_homelessness_cnt)) %>% 
  filter(experiencing_homelessness_cnt!="NA") %>%
  group_by(race,location) %>% 
  summarize(mean_count=mean(experiencing_homelessness_cnt)) %>% 
  arrange(-mean_count) %>% 
  ggplot(aes(x=reorder(race, mean_count),y=mean_count))+
  geom_col(color="black", fill="lightblue", alpha=0.75)+
  facet_wrap(.~location)+
  coord_flip()+
  labs(title = "Average Homeless Individuals-Yolo County CoC-Race",
       x="Race",
       y="Average Individuals that Recieved Care")
  

```

### Attempt at filling by race for every county
```{r}
race_homeless %>% 
  mutate(calendar_year=as.factor(calendar_year)) %>%
  mutate(experiencing_homelessness_cnt=as.numeric(experiencing_homelessness_cnt)) %>% 
  filter(experiencing_homelessness_cnt!="NA") %>%
  filter(location!="California") %>% 
  filter(calendar_year=="2023") %>% 
  filter(location!="Los Angeles County CoC") %>% 
  group_by(location, race) %>% 
  select(race, location, experiencing_homelessness_cnt) %>% 
  summarize(mean_count=mean(experiencing_homelessness_cnt)) %>% 
  ggplot(aes(x=location, y=mean_count, fill=race))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
  
```
### Top five counties with the highest rates of homelessness
```{r}
race_homeless %>% 
  mutate(calendar_year=as.factor(calendar_year)) %>%
  mutate(experiencing_homelessness_cnt=as.numeric(experiencing_homelessness_cnt)) %>% 
  filter(experiencing_homelessness_cnt!="NA") %>%
  filter(location!="California") %>% 
  filter(calendar_year=="2023") %>% 
  filter(location!="Los Angeles County CoC") %>% 
  group_by(location) %>% 
  summarise(total_count=sum(experiencing_homelessness_cnt)) %>% 
  top_n(5, total_count) %>% 
  arrange(-total_count)
```

```{r}
# Define custom fill colors
custom_colors <- c("powder blue","light blue","sky blue","gray","steel blue","dark blue", "blue")

race_homeless %>%
  mutate(calendar_year = as.factor(calendar_year),
         experiencing_homelessness_cnt = as.numeric(experiencing_homelessness_cnt)) %>%
  filter(!is.na(experiencing_homelessness_cnt),
         location != "California",
         calendar_year == "2023",
         location != "Los Angeles County CoC",
         location %in% c("San Diego County CoC", "Orange County CoC", "Santa Clara County CoC", "San Francisco CoC", "Fresno, Madera Counties CoC")) %>%
  group_by(location, race) %>%
  summarize(mean_count = mean(experiencing_homelessness_cnt)) %>%
  ggplot(aes(x = location, y = mean_count, fill = race)) +
  geom_col() +
  scale_fill_manual(values = custom_colors) +  # Apply custom fill colors
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "Average Homelessness Count by Location and Race in 2023",
       x="County",
       y="Average Count")  # Update plot title
```
### App for Race
```{r}
race_app <- race_homeless %>% 
  mutate(experiencing_homelessness_cnt=as.numeric(experiencing_homelessness_cnt)) %>% 
  filter(experiencing_homelessness_cnt!="NA"|experiencing_homelessness_cnt!="*") %>% 
  group_by(race, location, calendar_year) %>% 
  summarize(total_count=sum(experiencing_homelessness_cnt),
            average=mean(experiencing_homelessness_cnt))
```



```{r}


# Define UI
ui <- dashboardPage(
  dashboardHeader(title = "Location and Race Breakdown 2017-2023"),
  dashboardSidebar(
    selectInput("race", "Select Race", choices = unique(race_app$race)),
    selectInput("location", "Select Location", choices = unique(race_app$location)),
    selectInput("y_axis", "Select Y-axis", choices = c("Total Count", "Average"))
  ),
  dashboardBody(
    fluidRow(
      box(plotOutput("barPlot"), width = 12)
    )
  )
)

# Define server
server <- function(input, output) {
  
  # Filter data based on user input
  filtered_data <- reactive({
    race_app %>%
      mutate(calendar_year=as.factor(calendar_year)) %>%
      filter(race == input$race, location == input$location)
  })
  
  # Create bar plot
  output$barPlot <- renderPlot({
    ggplot(filtered_data(), aes(x = calendar_year, y = if (input$y_axis == "Total Count") total_count else average))+
      geom_col(color="black", fill="steelblue", alpha=0.75) +
      labs(title = paste(input$y_axis, "for", input$race, "in", input$location),
           x = "Year",
           y = input$y_axis)
  })
}

# Run the app
shinyApp(ui, server)
```

### Age App
```{r}
age_homeless <- clean_names(age_grouped)
```
```{r}
age_app <- age_homeless %>% 
  mutate(experiencing_homelessness_cnt=as.numeric(experiencing_homelessness_cnt)) %>% 
  filter(experiencing_homelessness_cnt!="NA"|experiencing_homelessness_cnt!="*") %>% 
  group_by(age_group, location, calendar_year) %>% 
  summarize(total_count=sum(experiencing_homelessness_cnt),
            average=mean(experiencing_homelessness_cnt))
```
```{r}
ui <- dashboardPage(
  dashboardHeader(title = "Age Group Breakdown 2017-2023"),
  dashboardSidebar(
    selectInput("age_group", "Select Age Group", choices = unique(age_app$age_group)),
    selectInput("location", "Select Location", choices = unique(age_app$location)),
    selectInput("y_axis", "Select Y-axis", choices = c("Total Count", "Average"))
  ),
  dashboardBody(
    fluidRow(
      box(plotOutput("barPlot"), width = 12)
    )
  )
)

# Define server
server <- function(input, output) {
  
  # Filter data based on user input
  filtered_data <- reactive({
    age_app %>%
      filter(age_group == input$age_group, location == input$location)
  })
  
  # Create bar plot
  output$barPlot <- renderPlot({
    ggplot(filtered_data(), aes(x = calendar_year, y = if (input$y_axis == "Total Count") total_count else average)) +
      geom_col(color = "black", fill = "steelblue", alpha = 0.75) +
      labs(title = paste(input$y_axis, "for Age Group", input$age_group, "in", input$location),
           x = "Year",
           y = input$y_axis)
  })
}

# Run the app
shinyApp(ui, server)
```

###Combined the two apps
```{r}
ui <- dashboardPage(
  dashboardHeader(title = "Location-Race-Age Breakdown 2017-2023"),
  dashboardSidebar(
    sidebarMenu(
      menuItem("Race", tabName = "race"),
      menuItem("Age", tabName = "age")
    )
  ),
  dashboardBody(
    tabItems(
      tabItem(tabName = "race",
              selectInput("race", "Select Race", choices = unique(race_app$race)),
              selectInput("location", "Select Location", choices = unique(race_app$location)),
              selectInput("y_axis", "Select Y-axis", choices = c("Total Count", "Average")),
              fluidRow(
                box(plotOutput("race_barPlot"), width = 12)
              )),
      tabItem(tabName = "age",
              selectInput("age_group", "Select Age Group", choices = unique(age_app$age_group)),
              selectInput("location_age", "Select Location", choices = unique(age_app$location)),
              selectInput("y_axis_age", "Select Y-axis", choices = c("Total Count", "Average")),
              fluidRow(
                box(plotOutput("age_barPlot"), width = 12)
              ))
    )
  )
)

# Define server
server <- function(input, output) {
  
  # Filter data based on user input (race)
  filtered_race_data <- reactive({
    race_app %>%
      filter(race == input$race, location == input$location)
  })
  
  # Filter data based on user input (age)
  filtered_age_data <- reactive({
    age_app %>%
      filter(age_group == input$age_group, location == input$location_age)
  })
  
  # Create bar plot for race app
  output$race_barPlot <- renderPlot({
    ggplot(filtered_race_data(), aes(x = calendar_year, y = if (input$y_axis == "Total Count") total_count else average)) +
      geom_col(color = "black", fill = "steelblue", alpha = 0.75) +
      labs(title = paste(input$y_axis, "for", input$race, "in", input$location),
           x = "Year",
           y = input$y_axis)
  })
  
  # Create bar plot for age app
  output$age_barPlot <- renderPlot({
    ggplot(filtered_age_data(), aes(x = calendar_year, y = if (input$y_axis_age == "Total Count") total_count else average)) +
      geom_col(color = "black", fill = "steelblue", alpha = 0.75) +
      labs(title = paste(input$y_axis_age, "for Age Group", input$age_group, "in", input$location_age),
           x = "Year",
           y = input$y_axis_age)
  })
}

# Run the app
shinyApp(ui, server)
```

