---
title: "Homeless Project"
output: html_document
date: "2024-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
install.packages("shinydashboard")
```

```{r}
library(tidyverse)
library(dplyr)
library(naniar)
library(ggplot2)
library(shiny)
library(shinydashboard)
```



```{r}
age <- read_csv("age.csv")
race <- read_csv("race.csv")
```

```{r}
glimpse(age)
glimpse(race)
```

```{r}
miss_var_summary(race)
```


#Getting an idea of total counts of individuals experiencing homelessness in each location every year recorded
```{r}
age %>% 
  group_by(CALENDAR_YEAR, LOCATION) %>% 
  mutate(EXPERIENCING_HOMELESSNESS_CNT=as.integer(EXPERIENCING_HOMELESSNESS_CNT)) %>% 
  summarise(total_homeless=sum(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE)) %>% 
  arrange(desc(total_homeless))
```

#Identifying hotspots by county for homeless individuals across all years recorded
```{r}
age %>% 
  group_by(LOCATION) %>% 
  filter(LOCATION!="California") %>% 
  mutate(EXPERIENCING_HOMELESSNESS_CNT=as.integer(EXPERIENCING_HOMELESSNESS_CNT)) %>% 
  summarise(total_homeless=sum(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE)) %>% 
  arrange(desc(total_homeless))
```

#Identify homelessness by year (noting that individuals increased after COVID in 2020)
```{r}
age %>% 
  group_by(CALENDAR_YEAR) %>% 
  mutate(EXPERIENCING_HOMELESSNESS_CNT=as.integer(EXPERIENCING_HOMELESSNESS_CNT)) %>% 
  summarise(total_homeless=sum(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE)) %>% 
  arrange(desc(total_homeless))
```

#Total homeless count over all years
```{r}
age %>% 
  group_by(CALENDAR_YEAR) %>% 
  mutate(CALENDAR_YEAR=as.factor(CALENDAR_YEAR)) %>% 
  mutate(EXPERIENCING_HOMELESSNESS_CNT=as.integer(EXPERIENCING_HOMELESSNESS_CNT)) %>% 
  summarise(total_homeless=sum(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE)) %>% 
  ggplot(aes(x=CALENDAR_YEAR, y=total_homeless))+ 
  geom_col(color="black", fill="blue", alpha=0.6)+
  labs(title = "Total Homeless Count 2017-2023",
       x= "Year",
       y= "Total Count")
```
```{r}
class(age$AGE_GROUP_PUBLIC)
table(age$AGE_GROUP_PUBLIC)
```

#Use case_when to create new categories with uniform data
```{r}
age_grouped <- age %>% 
  filter(AGE_GROUP_PUBLIC!="Unknown") %>% 
  mutate(age_group = case_when(
    AGE_GROUP_PUBLIC == "Under 18" ~ "<18",
    AGE_GROUP_PUBLIC == "18-24" ~ "18-24",
    AGE_GROUP_PUBLIC == "25-34" ~ "25-34",
    AGE_GROUP_PUBLIC == "35-44" ~ "35-44",
    AGE_GROUP_PUBLIC == "45-54" ~ "45-54",
    AGE_GROUP_PUBLIC == "55-64" ~ "55-64",
    AGE_GROUP_PUBLIC == "65+" ~ ">65")) %>% 
  filter(!is.na(age_group))
head(age_grouped)
```

#summarise total homelessness by age group
```{r}
age_grouped %>% 
  group_by(age_group) %>% 
  mutate(EXPERIENCING_HOMELESSNESS_CNT=as.integer(EXPERIENCING_HOMELESSNESS_CNT)) %>% 
  summarise(total_homeless=sum(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE)) %>% 
  mutate(age_group = factor(age_group, levels = c("<18", "18-24", "25-34", "35-44", "45-54", "55-64", ">65"))) %>%
  arrange(desc(total_homeless))
```


# plot it and get rid of scientific notation
```{r}
age_grouped %>% 
  group_by(age_group) %>% 
  mutate(EXPERIENCING_HOMELESSNESS_CNT=as.integer(EXPERIENCING_HOMELESSNESS_CNT)) %>% 
  summarise(total_homeless=sum(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE)) %>% 
  mutate(age_group = factor(age_group, levels = c("<18", "18-24", "25-34", "35-44", "45-54", "55-64", ">65"))) %>%
  ggplot(aes(x=age_group, y=total_homeless))+
  geom_col(color="black", fill="lightblue", alpha=0.75)+
  labs(title = "Total Homelessness by Age Group",
       x= "Age Group",
       y= "Total Count")+
  scale_y_continuous(labels = scales::comma)
```
# Plotting average homelessness in age groups by county (obviously too many to look at)... Discovered we want to make a shiny app to visualize this analysis
```{r}
age_grouped %>% 
  filter(LOCATION!="California") %>% 
  group_by(age_group, LOCATION) %>% 
  mutate(EXPERIENCING_HOMELESSNESS_CNT=replace_na(as.integer(EXPERIENCING_HOMELESSNESS_CNT), 0)) %>% 
  summarise(average_homeless=mean(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE)) %>% 
  arrange(desc(average_homeless)) %>% 
  ggplot(aes(x=age_group, y=average_homeless))+
  geom_col()+
  facet_wrap(~LOCATION, scales = "free")
```


#Plot Average Homeless Individuals in Yolo County Coc by age group
```{r}
age_grouped %>% 
  filter(LOCATION == "Yolo County CoC") %>% 
  group_by(age_group) %>% 
  mutate(EXPERIENCING_HOMELESSNESS_CNT = replace_na(as.integer(EXPERIENCING_HOMELESSNESS_CNT), 0)) %>%
  summarise(mean_homeless = mean(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE)) %>%
  mutate(age_group = factor(age_group, levels = c("<18", "18-24", "25-34", "35-44", "45-54", "55-64", ">65"))) %>%
  ggplot(aes(x = age_group, y = mean_homeless)) +
  geom_col(color = "black", fill = "lightblue", alpha = 0.75) +
  coord_flip() +
  labs(title = "Average Homeless Individuals in Yolo County Coc-Age Group",
       x = "Age Group",
       y = "Average Individuals that Received Care")
```


# What age group existing in which county have the most total homeless population and what is the average?
```{r}
age_grouped %>% 
  filter(LOCATION!="California") %>% 
  group_by(age_group, LOCATION) %>% 
  mutate(EXPERIENCING_HOMELESSNESS_CNT=as.integer(EXPERIENCING_HOMELESSNESS_CNT)) %>% 
  summarise(total_homeless=sum(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE),
            average_homeless=mean(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE)) %>% 
  arrange(desc(total_homeless))
```
#create new object for 
```{r}
app <- age_grouped %>% 
  filter(LOCATION!="California") %>% 
  mutate(EXPERIENCING_HOMELESSNESS_CNT = replace_na(as.integer(EXPERIENCING_HOMELESSNESS_CNT), 0)) %>% 
  filter(EXPERIENCING_HOMELESSNESS_CNT!="NA")
```


#create new variable defining pre covid and covid affected to compare
```{r}
covid <- app %>% 
  mutate(CALENDAR_YEAR=as.numeric(CALENDAR_YEAR)) %>% 
  mutate(Covid = case_when(
    CALENDAR_YEAR <= 2019 ~ "Pre-Covid",
    CALENDAR_YEAR >= 2020 ~ "Covid"
  ))

```


#Plot a hotspot county of choice to compare pre covid and post covid average homelessness
```{r}
covid %>% 
  group_by(Covid) %>% 
  filter(LOCATION=="Los Angeles County CoC") %>% 
  summarise(mean_homeless=mean(EXPERIENCING_HOMELESSNESS_CNT)) %>% 
  ggplot(aes(x=Covid, y=mean_homeless))+
  geom_col(color="black", fill="lightblue", alpha=0.75)+
  labs(title = "Average Homelessness Before and During COVID-19 in Los Angeles County CoC",
       x= "Covid Status",
       y= "Average Homelessness")
```

#experiment with shiny apps
```{r}
ui <- dashboardPage(
  dashboardHeader("Age Group by Total Homeless Count Compared by County"),
  dashboardSidebar(
    selectInput("Age", "Select Age Group", choices = c("<18", "18-24", "25-34", "35-44", "45-54", "55-64", ">65"), selected = "<18"),
    selectInput("Statistics", "Select Total or Mean", choices = c("Total Count", "Average Count"), selected = "Total Count"),
    selectInput("County", "Select County", choices = unique(app$LOCATION))
  ),
  dashboardBody(
    fluidRow(
      box(plotOutput("barPlot"), width = 12))
    )
  )
)

server <- function(input, output, session) {
  filtered_data <- reactive({
    filtered <- app %>%
      filter(LOCATION == input$County) %>%
      filter(age_group == input$Age) %>%
      summarise(total_homeless = sum(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE),
                average_homeless = mean(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE))
    
  })
  
  output$plot <- renderPlot({
    ggplot(filtered_data(), aes(x = CALENDAR_YEAR, y = if (input$Statistics == "Total Count") total_homeless else average_homeless))+
      geom_col(color = "black", fill = "lightblue", alpha = 0.75) +
      labs(title = paste(input$y_axis, "for", input$Age, "in", input$COUNTY),
           x = "Year",
           y = input$Statistics)
    
  })
}

shinyApp(ui, server)


```


```{r}
library(shiny)

ui <- fluidPage(
  titlePanel("Age Group by Total Homeless Count Compared by County"),
  selectInput("Age","Select Age Group", choices = c("<18","18-24","25-34","35-44","45-54","55-64",">65"),
                            selected = "<18"),
  selectInput("Statistics", "Select Total or Mean", choices = c("Total Count","Average Count"),
                            selected = "Total Count"),
  selectInput("County", "Select County", choices = unique(app$LOCATION)),
  plotOutput("plot", width = "500px", height = "400px")
                            )
  


server <- function(input, output, session) {
  filtered_data <- reactive({
    filtered <- app %>% 
      mutate(CALENDAR_YEAR=as.factor(CALENDAR_YEAR)) %>%
      filter(LOCATION == input$County) %>%
      filter(age_group == input$age_group) %>%
      summarise(total_homeless=sum(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE),
                average_homeless=mean(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE))
    
    if (input$Statistics == "Total Count") {
      filtered <- filtered %>%
        summarise(total_homeless = sum(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE))
    } else {
      filtered <- filtered %>%
        summarise(average_homeless = mean(EXPERIENCING_HOMELESSNESS_CNT, na.rm = TRUE))
    }
    
    filtered
  })
     
    
    output$plot <- renderPlot({
      ggplot(filtered_data(), aes(x=age_group, y= ifelse(input$Statistics == "Total Count", total_homeless, average_homeless)))+
        geom_col(color="black", fill="lightblue", alpha=0.75)
  
})
}
shinyApp(ui, server)
```






